---
output: github_document
---

<!-- 
README.md is generated from README.Rmd. Please edit that file
rmarkdown::render("README.Rmd") 
-->


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/lognorm)](http://cran.r-project.org/package=lognorm)
[![Travis-CI Build Status](https://travis-ci.org/bgctw/lognorm.svg?branch=master)](https://travis-ci.org/bgctw/lognorm)


## Overview

`lognorm` package provides support for the univariate
[logit-normal
distribution](https://en.wikipedia.org/wiki/Logit-normal_distribution). In
addition to the usual random, density, percential, and quantile function, it
helps with estimating distribution parameters from observations statistics.

## Installation

```{r, eval = FALSE}
# From CRAN
install.packages("lognorm")

# Or the the development version from GitHub:
# install.packages("devtools")
devtools::install_github("bgctw/lognorm")
```

## Usage

See the package vignettes for an introduction.

A simple example computes the distribution parameters of the sum of 
two correlated lognormal parameters.
 
```{r example}
means <- c(110,100)
sigmaStar <- c(1.5,1.5)
corr <- setMatrixOffDiagonals(diag(nrow = 2), value = 0.6, isSymmetric = TRUE)
coefTerms <- getParmsLognormForExpval(means, sigmaStar)
#estimateSumLognormalBenchmark( c(mu1,mu2), c(sigma1,sigma2) )
coefSum <- estimateSumLognormal( coefTerms[,"mu"], coefTerms[,"sigma"], corr = corr )
# 
x <- seq(50,500,length.out = 100)
density <- dlnorm(x, coefSum["mu"], coefSum["sigma"])
plot(density ~ x, type = "l")
# confidence intervals
abline(v = qlnorm(c(0.025, 0.975), coefSum["mu"], coefSum["sigma"]), lty = "dotted")
# expected value
abline(v = getLognormMoments(coefSum["mu"], coefSum["sigma"])[1,"mean"])
# mode
abline(v = getLognormMode(coefSum["mu"], coefSum["sigma"]), lty = "dashed")
# median
abline(v = getLognormMedian(coefSum["mu"], coefSum["sigma"]), lty = "dotdash")
```

The sum of the expected values is conserved, while the mutliplicative standard 
deviation decreases during aggregtion:

```{r}
c( 
  mean = unname(getLognormMoments(coefSum["mu"], coefSum["sigma"])[1,"mean"])
  , sigmaStar = unname(exp(coefSum["sigma"])))
```

