---
output: 
  rmarkdown::html_vignette: 
    keep_md: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Approximating the difference of two lognormal random variables}
  %\usepackage[UTF-8]{inputenc}
---
```{r eval=FALSE, include=FALSE}
# twDev::genVigs()
rmarkdown::render("lognormalDiff.Rmd","md_document") 
```

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra = 'style="display:block; margin: auto"'
    #, fig.align = "center"
    #, fig.width = 4.6, fig.height = 3.2
    , fig.width = 6, fig.height = 3.75 #goldener Schnitt 1.6
    , dev.args = list(pointsize = 10)
    , dev = c('png','pdf')
    )
knit_hooks$set(spar = function(before, options, envir) {
    if (before) {
        par( las = 1 )                   #also y axis labels horizontal
        par(mar = c(2.0,3.3,0,0) + 0.3 )  #margins
        par(tck = 0.02 )                          #axe-tick length inside plots             
        par(mgp = c(1.1,0.2,0) )  #positioning of axis title, axis labels, axis
     }
})
library(lognorm) 
if (!require(ggplot2) || !require(dplyr) || !require(tidyr) || !require(purrr)) {
	print("To generate this vignette, ggplo2, dplyr, tidyr, and purrr are required.")
	exit(0)
}
themeTw <- theme_bw(base_size = 10) + 
  theme(axis.title = element_text(size = 9))
```

# Approximating the difference of lognormal random variables

Lo 2012 reports an approximation for the difference of two random variables by
a shifted lognormal distribution.

Rather than approximating the density of $y = a - b$, 
it approximates the density of $y_s = a - b + s$, where
$s$ is the shift. 
Hence, one has to subtract $s$ from provided mean and quantiles. One can
can use the variance, and relative error but has to recompute the relative error.

## Two uncorrelated random variables
```{r}
# generate nSample values of two lognormal random variables
mu1 = log(110)
mu2 = log(100)
sigma1 = 0.25
sigma2 = 0.15
#(coefSum <- estimateSumLognormal( c(mu1,mu2), c(sigma1,sigma2) ))
(coefSum <- estimateDiffLognormal(mu1,mu2, sigma1,sigma2, 0))
(expSum <- getLognormMoments(coefSum["mu"], coefSum["sigma"])[,"mean"] - 
    coefSum["shift"])
```

Several functions accept the `shift` argument to handle this already.
```{r}
getLognormMoments(coefSum["mu"], coefSum["sigma"], shift = coefSum["shift"])
getLognormMode(coefSum["mu"], coefSum["sigma"], shift = coefSum["shift"])
getLognormMedian(coefSum["mu"], coefSum["sigma"], shift = coefSum["shift"])
```

For the functions from the stats package, the shifting has to be done
manually.
```{r}
p <- seq(0,1,length.out = 100)[-c(1,100)]
dsPredY <- data.frame(
  var = "y", 
  q_shifted = qlnorm(p, coefSum["mu"], coefSum["sigma"] )
) %>%
mutate( 
  q = q_shifted - coefSum["shift"],
  d = dlnorm(q_shifted, coefSum["mu"], coefSum["sigma"])
)
```

A check by random numbers (dotted lines) shows close correspondence.

```{R plotUncorr, echo=FALSE, fig.height=2.04, fig.width=3.27}
nSample = 2000

ds <- data.frame(
  a = rlnorm(nSample, mu1, sigma1)
  , b = rlnorm(nSample, mu2, sigma2)
) %>%  mutate(
  y = a - b
)
dsw <- gather(ds, "var", "value", a, b, y)
p1 <- dsw %>% filter(var == "y") %>% 
  ggplot(aes(value, color = var)) + geom_density(linetype = "dotted") +
  geom_vline(xintercept = mean(ds$y), linetype = "dotted")
p1 + geom_line(data = dsPredY, aes(q, d, color = var)) +
  geom_vline(aes(xintercept = expSum, color = var), 
             data = data.frame(var = "y", q = expSum)) +
themeTw +
theme(legend.position = c(0.98,0.98), legend.justification = c(1,1)) +
theme(axis.title.x = element_blank())
```
## Test if difference is significantly different from zero

The probability of the zero quantile needs to larger than a significane 
level.
We can compute it based on the lognormal approximation or by repeatetly
computing the difference from a sample of terms.

```{r}
  mu1 = log(120)
  mu2 = log(60)
  sigma1 = 0.25
  sigma2 = 0.15
  coefSum <- estimateDiffLognormal( mu1,mu2,sigma1,sigma2, corr = -0.8 )
  pLo <- plnorm(0 + coefSum["shift"], coefSum["mu"], coefSum["sigma"])
  pSample <- pDiffLognormalSample(mu1,mu2,sigma1,sigma2, corr = -0.8)
  c(pLo = as.numeric(pLo), pSample = pSample)
```

In the example both approaches give a probability of less than 5% so
that we conclude that the difference is significant.

```{r plotSignifZero, echo=FALSE}
p <- seq(0,1,length.out = 100)[-c(1,100)]
dsPredY <- tibble(
  var = "y", 
  q_shifted = qlnorm(p, coefSum["mu"], coefSum["sigma"] ),
  q = q_shifted - coefSum["shift"],
  d = dlnorm(q_shifted, coefSum["mu"], coefSum["sigma"])
)
dsPredA <- tibble(
  var = "a", 
  q = qlnorm(p, mu1, sigma1),
  d = dlnorm(q, mu1, sigma1)
)
dsPredB <- tibble(
  var = "b", 
  q = qlnorm(p, mu2, sigma2),
  d = dlnorm(q, mu2, sigma2)
)
bind_rows(select(dsPredY, -q_shifted), dsPredA, dsPredB) %>% 
  #filter(var == "a") %>% 
  ggplot(aes(q, d, color = var)) + geom_line() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkgrey") +
themeTw +
theme(legend.position = c(0.98,0.98), legend.justification = c(1,1)) +
theme(axis.title.x = element_blank())
```

## Two positively correlated variables
```{r}
if (!requireNamespace("mvtnorm")) {
  warning("Remainder of the vignette required mvtnorm installed.")
  knitr::opts_chunk$set(error = TRUE) 
}
corr = 0.8
(coefSum <- estimateDiffLognormal(mu1,mu2, sigma1,sigma2, corr = corr))
(expSum <- getLognormMoments(coefSum["mu"], coefSum["sigma"])[,"mean"] - 
    coefSum["shift"])
```

Check with sampled distribution.
```{r}
nSample <- 1e5
sigma_vec = c(sigma1, sigma2)
corrM <- setMatrixOffDiagonals(
  diag(nrow = 2), value = corr, isSymmetric = TRUE)
covM <- diag(sigma_vec) %*% corrM %*% diag(sigma_vec)
xObsN <- exp(mvtnorm::rmvnorm(nSample, mean =  c(mu1, mu2), sigma = covM))
head(xObsN)
y = xObsN[,1] - xObsN[,2]
```

```{r pdfDiffCorr80, echo=FALSE, fig.height=2.04, fig.width=3.27}
p <- seq(0,1,length.out = 100)[-c(1,100)]
dsPredY <- data.frame(
  var = "y", 
  q_shifted = qlnorm(p, coefSum["mu"], coefSum["sigma"] )
) %>%
  mutate( 
    q = q_shifted - coefSum["shift"],
    d = dlnorm(q_shifted, coefSum["mu"], coefSum["sigma"])
    )
# density plot of the random draws
ggplot(data.frame(y = y), aes(y, color = "random draws")) + 
  geom_density() +
  # line plot of the lognorm density approximation
  geom_line(data = dsPredY, aes(q, d, color = "computed diff")) +
  # expected value
  geom_vline(xintercept = expSum, linetype = "dashed") +
themeTw +
theme(legend.position = c(0.98,0.98), legend.justification = c(1,1)) +
theme(axis.title.x = element_blank()) +
theme(legend.title = element_blank())
```

The approximation for the difference of negatively correlated random numbers 
correcly predicts a 
narrower distribution than with the uncorrelated difference
but shows some deviations from the sampled distribution  around the mode.

## Subtracting a variable with larger variance

The method only works if the $\sigma_b < \sigma_a$ and otherwise
gives an error.

```{r}
# generate nSample values of two lognormal random variables
mu1 = log(110)
mu2 = log(100)
sigma1 = 0.15
sigma2 = 0.25
#(coefSum <- estimateSumLognormal( c(mu1,mu2), c(sigma1,sigma2) ))
try(coefSum <- estimateDiffLognormal(mu1,mu2, sigma1,sigma2, 0))
```

But one can compute the density of $y_r = -y = b - a$ and plot 
the density of the shifted and negated distribution.

```{r}
# generate nSample values of two lognormal random variables
mu1 = log(110)
mu2 = log(100)
sigma1 = 0.15
sigma2 = 0.25
#(coefSum <- estimateSumLognormal( c(mu1,mu2), c(sigma1,sigma2) ))
(coefSum <- estimateDiffLognormal(mu2,mu1, sigma2,sigma1, 0))
(expSum <- -(getLognormMoments(coefSum["mu"], coefSum["sigma"])[,"mean"] - 
    coefSum["shift"]))
```
```{r}
p <- seq(0,1,length.out = 100)[-c(1,100)]
dsPredY <- data.frame(
  var = "y", 
  q_shifted_neg = qlnorm(p, coefSum["mu"], coefSum["sigma"] )
) %>%
  mutate( 
    q = -(q_shifted_neg - coefSum["shift"]),
    d = dlnorm(q_shifted_neg, coefSum["mu"], coefSum["sigma"])
    )
```

```{r plotDiffLargerVar, echo=FALSE, fig.height=2.04, fig.width=3.27}
nSample = 2000
ds <- data.frame(
  a = rlnorm(nSample, mu1, sigma1)
  , b = rlnorm(nSample, mu2, sigma2)
) %>%  mutate(
  y = a - b
)
dsw <- gather(ds, "var", "value", a, b, y)
p1 <- dsw %>% filter(var == "y") %>% 
  ggplot(aes(value, color = var)) + geom_density(linetype = "dotted") +
  geom_vline(xintercept = mean(ds$y), linetype = "dotted")
#

p1 + geom_line(data = dsPredY, aes(q, d, color = var)) +
  geom_vline(aes(xintercept = expSum, color = var), 
             data = data.frame(var = "y", q = expSum)) +
themeTw +
theme(legend.position = c(0.98,0.98), legend.justification = c(1,1)) +
theme(axis.title.x = element_blank())
```

