---
output: 
  rmarkdown::html_vignette: 
    keep_md: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Approximating the difference of two lognormal random variables}
  %\usepackage[UTF-8]{inputenc}
---
```{r eval=FALSE, include=FALSE}
# twDev::genVigs()
#rmarkdown::render("lognormalDiff.Rmd","md_document") 
```

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra = 'style="display:block; margin: auto"'
    #, fig.align = "center"
    #, fig.width = 4.6, fig.height = 3.2
    , fig.width = 6, fig.height = 3.75 #goldener Schnitt 1.6
    , dev.args = list(pointsize = 10)
    , dev = c('png','pdf')
    )
knit_hooks$set(spar = function(before, options, envir) {
    if (before) {
        par( las = 1 )                   #also y axis labels horizontal
        par(mar = c(2.0,3.3,0,0) + 0.3 )  #margins
        par(tck = 0.02 )                          #axe-tick length inside plots             
        par(mgp = c(1.1,0.2,0) )  #positioning of axis title, axis labels, axis
     }
})
library(lognorm) 
if (!require(ggplot2) || !require(dplyr) || !require(tidyr) || !require(purrr)) {
	print("To generate this vignette, ggplot2, dplyr, tidyr, and purrr are required.")
	knit_exit()
}
themeTw <- ggplot2::theme_bw(base_size = 10) + 
  theme(axis.title = element_text(size = 9))
```

# Approximating the difference of lognormal random variables

Lo 2012 reports an approximation for the difference of two random variables by
a shifted lognormal distribution.

Rather than approximating the density of $y = a - b$, 
it approximates the density of $y_s = a - b + s$, where
$s$ is the shift. 
Hence, one has to subtract $s$ from provided mean and quantiles. One can
can use the variance, and relative error but has to recompute the relative error.

## Two uncorrelated random variables
```{r}
# generate nSample values of two lognormal random variables
mu1 = log(110)
mu2 = log(100)
sigma1 = 0.25
sigma2 = 0.15
#(coefDiff <- estimateSumLognormal( c(mu1,mu2), c(sigma1,sigma2) ))
(coefDiff <- estimateDiffLognormal(mu1,mu2, sigma1,sigma2, 0))
(expDiff <- getLognormMoments(coefDiff["mu"], coefDiff["sigma"])[,"mean"] - 
    coefDiff["shift"])
```

Several functions accept the `shift` argument to handle this already.
```{r}
getLognormMoments(coefDiff["mu"], coefDiff["sigma"], shift = coefDiff["shift"])
getLognormMode(coefDiff["mu"], coefDiff["sigma"], shift = coefDiff["shift"])
getLognormMedian(coefDiff["mu"], coefDiff["sigma"], shift = coefDiff["shift"])
```

For the functions from the stats package, the shifting has to be done
manually.
```{r}
p <- seq(0,1,length.out = 100)[-c(1,100)]
dsPredY <- data.frame(
  var = "y", 
  q_shifted = qlnorm(p, coefDiff["mu"], coefDiff["sigma"] )
) %>%
mutate( 
  q = q_shifted - coefDiff["shift"],
  d = dlnorm(q_shifted, coefDiff["mu"], coefDiff["sigma"])
)
```

A check by random numbers (dotted lines) shows close correspondence.

```{R plotUncorr, echo=FALSE, fig.height=2.04, fig.width=3.27}
nSample = 2000

ds <- data.frame(
  a = rlnorm(nSample, mu1, sigma1)
  , b = rlnorm(nSample, mu2, sigma2)
) %>%  mutate(
  y = a - b
)
dsw <- gather(ds, "var", "value", a, b, y)
p1 <- dsw %>% filter(var == "y") %>% 
  ggplot(aes(value, color = var)) + geom_density(linetype = "dotted") +
  geom_vline(xintercept = mean(ds$y), linetype = "dotted")
p1 + geom_line(data = dsPredY, aes(q, d, color = var)) +
  geom_vline(aes(xintercept = expDiff, color = var), 
             data = data.frame(var = "y", q = expDiff)) +
themeTw +
theme(legend.position = c(0.98,0.98), legend.justification = c(1,1)) +
theme(axis.title.x = element_blank())
```

## Test if difference is significantly different from zero

The probability of the zero quantile needs to be larger than a significance 
level.
We can compute it based on the lognormal approximation.

Since Lo12 is only accurate if the expected difference is small compared
to the expected sum, the probability of the difference being larger than zero can be estimated by a sampling both terms.

```{r}
  mu1 = log(120)
  mu2 = log(60)
  sigma1 = 0.25
  sigma2 = 0.15
  coefDiff <- estimateDiffLognormal( mu1,mu2,sigma1,sigma2, corr = -0.8 )
  pLo <- plnorm(0 + coefDiff["shift"], coefDiff["mu"], coefDiff["sigma"])
  pSample <- pDiffLognormalSample(mu1,mu2,sigma1,sigma2, corr = -0.8)
  c(pLo = as.numeric(pLo), pSample = pSample)
```

In the example both approaches give a probability of less than 5% so
that we conclude that the difference is significant.

```{r plotSignifZero, echo=FALSE}
p <- seq(0,1,length.out = 100)[-c(1,100)]
dsPredY <- tibble(
  var = "y", 
  q_shifted = qlnorm(p, coefDiff["mu"], coefDiff["sigma"] ),
  q = q_shifted - coefDiff["shift"],
  d = dlnorm(q_shifted, coefDiff["mu"], coefDiff["sigma"])
)
dsPredA <- tibble(
  var = "a", 
  q = qlnorm(p, mu1, sigma1),
  d = dlnorm(q, mu1, sigma1)
)
dsPredB <- tibble(
  var = "b", 
  q = qlnorm(p, mu2, sigma2),
  d = dlnorm(q, mu2, sigma2)
)
bind_rows(select(dsPredY, -q_shifted), dsPredA, dsPredB) %>% 
  #filter(var == "a") %>% 
  ggplot(aes(q, d, color = var)) + geom_line() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkgrey") +
themeTw +
theme(legend.position = c(0.98,0.98), legend.justification = c(1,1)) +
theme(axis.title.x = element_blank())
```

## Two positively correlated variables
```{r}
if (!requireNamespace("mvtnorm")) {
  warning("Remainder of the vignette required mvtnorm installed.")
  knitr::opts_chunk$set(error = TRUE) 
}
corr = 0.8
(coefDiff <- estimateDiffLognormal(mu1,mu2, sigma1,sigma2, corr = corr))
(expDiff <- getLognormMoments(
  coefDiff["mu"], coefDiff["sigma"], shift = coefDiff["shift"])[,"mean"])
```

Check with sampled distribution.
```{r}
nSample <- 1e5
sigma_vec = c(sigma1, sigma2)
corrM <- setMatrixOffDiagonals(
  diag(nrow = 2), value = corr, isSymmetric = TRUE)
covM <- diag(sigma_vec) %*% corrM %*% diag(sigma_vec)
xObsN <- exp(mvtnorm::rmvnorm(nSample, mean =  c(mu1, mu2), sigma = covM))
head(xObsN)
y = xObsN[,1] - xObsN[,2]
```

```{r pdfDiffCorr80, echo=FALSE, fig.height=2.04, fig.width=3.27}
p <- seq(0,1,length.out = 100)[-c(1,100)]
dsPredY <- data.frame(
  var = "y", 
  q_shifted = qlnorm(p, coefDiff["mu"], coefDiff["sigma"] )
) %>%
  mutate( 
    q = q_shifted - coefDiff["shift"],
    d = dlnorm(q_shifted, coefDiff["mu"], coefDiff["sigma"])
    )
# density plot of the random draws
ggplot(data.frame(y = y), aes(y, color = "random draws")) + 
  geom_density() +
  # line plot of the lognorm density approximation
  geom_line(data = dsPredY, aes(q, d, color = "computed diff")) +
  # expected value
  geom_vline(xintercept = expDiff, linetype = "dashed") +
themeTw +
theme(legend.position = c(0.98,0.98), legend.justification = c(1,1)) +
theme(axis.title.x = element_blank()) +
theme(legend.title = element_blank())
```

The approximation for the difference of positively correlated random numbers 
predicts a 
narrower distribution than with the uncorrelated or negatively correlated difference above. However, this case less accurate and shows some deviations from the sampled distribution around the mode.

## Subtracting a variable with larger variance

The method of Lo12 requires $\sigma_b < \sigma_a$ and otherwise
gives an error.

```{r}
# generate nSample values of two lognormal random variables
mu1 = log(110)
mu2 = log(100)
sigma1 = 0.15
sigma2 = 0.25
try(coefDiff <- estimateDiffLognormal(mu1,mu2, sigma1,sigma2, 0))
```

But one can compute the density of $y_r = -y = b - a$ and plot 
the density of the shifted and negated distribution.

```{r}
# note the switch of positions of mu1 and mu2: mu2 - mu1
#(coefDiff <- estimateDiffLognormal(mu1,mu2, sigma1,sigma2, 0)
(coefDiff <- estimateDiffLognormal(mu2,mu1, sigma2,sigma1, 0))
# note the minus sign in front
(expDiff <- -(getLognormMoments(
  coefDiff["mu"], coefDiff["sigma"], shift = coefDiff["shift"])[,"mean"]))
```
```{r}
p <- seq(0,1,length.out = 100)[-c(1,100)]
dsPredY <- data.frame(
  var = "y", 
  q_shifted_neg = qlnorm(p, coefDiff["mu"], coefDiff["sigma"] )
) %>%
  mutate( 
    # note the minus sign in front
    q = -(q_shifted_neg - coefDiff["shift"]),
    d = dlnorm(q_shifted_neg, coefDiff["mu"], coefDiff["sigma"])
    )
```

```{r plotDiffLargerVar, echo=FALSE, fig.height=2.04, fig.width=3.27}
nSample = 2000
ds <- data.frame(
  a = rlnorm(nSample, mu1, sigma1)
  , b = rlnorm(nSample, mu2, sigma2)
) %>%  mutate(
  y = a - b
)
dsw <- gather(ds, "var", "value", a, b, y)
p1 <- dsw %>% filter(var == "y") %>% 
  ggplot(aes(value, color = var)) + geom_density(linetype = "dotted") +
  geom_vline(xintercept = mean(ds$y), linetype = "dotted")
#

p1 + geom_line(data = dsPredY, aes(q, d, color = var)) +
  geom_vline(aes(xintercept = expDiff, color = var), 
             data = data.frame(var = "y", q = expDiff)) +
themeTw +
theme(legend.position = c(0.98,0.98), legend.justification = c(1,1)) +
theme(axis.title.x = element_blank())
```
Because we subtract a large-variance lognormal variable, the distribution
becomes right-skewed.

